Governing through patronage: the bargain for education in decentralized Brazil

### 2020-10-21 paper/paper.tex
reframe drafting

% what do I need to bring the audience onboard?
% emphasize the existence of discretion, turnover, political manipulation of public sresources.
% in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns

Across the world, vulnerable populations rely on governments for access to basic public services such as education. In the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}. The World Bank refers to this phenomenon as ``learning poverty'': stunted learning when access to a schooling exists.\footnote{For more information, see \hyperlink{here}{https://www.worldbank.org/en/topic/education/brief/learning-poverty}.} These outcomes result from political choices, but a growing body of literature recognizes the important role bureaucratic institutions mediating these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015}.

Understanding how these political decisions cascade down the bureaucracy remains a challenge. Public services -- and their quality -- are the end-result of a series of interlocking steps: from personnel decisions to service provision \citep{pepinsky_bureaucracy_2017}, each link in this institutional chain of command connects political principals to the child sitting in the classroom. This panorama of decisions belie a multiplicity of actors that is often obscured in prevailing models of

How can we intervene in this chain to improve the quality of public services? Previous studies have explored how politicians' incentives can improve the quality of public goods provision. Some have focused on the direct link between the political principal and bureaucratic agent: increasing political oversight over bureaucracies may improve public services \citep{gulzar_politicians_2017, raffler_does_2016}. Others have focused on the more indirect electoral accountability link binding politicians to voters, exploring how to strengthen it to induce positive changes \citep{boas_electoral_2019, besley_political_2003} Whether explicitly or implicitly, these theories posit a unitary political actor who interacts with either the bureaucracy or the electorate. Yet in much of the developing world, these politicians do not govern alone: in fact, they are embedded in democratic institutions where multiplicity of elected officials and power-sharing arrangements dramatically reshape the institutional chain connecting politician to public service provision \citep{francois_how_2015}.

To shed light on this complexity, over two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.

>>>>

% what do I need to bring the audience onboard?
% emphasize the existence of discretion, turnover, political manipulation of public resources.
% in particular the existence of power-sharing arrangements between elected officials that go beyond electoral concerns

Across the world, vulnerable populations rely on governments for access to basic public services such as education. Yet in the developing world, the quality of public education remains low: functional illiteracy is widespread, and basic mathematical operations beyond the reach of many \citep{vegas_raising_2007}. A growing recognition of the importance of bureaucrats in explaining these outcomes \citep{pepinsky_bureaucracy_2017,finan_personnel_2015} has been accompanied by concern over the politician-bureaucrat nexus \citet{raffler_does_2016,gulzar_politicians_2017}. These actors, political and bureaucratic, have been extensively modeled under the dyadic framework of principal-agent models \citep{miller2005political}. Yet in most developing countries, these political and bureaucratic actors are multiple and the institutional environment under which they operate play a decisive role in how public services are managed and delivered.

In developing democracies, and within their political constituencies, politicians are seldom unitary. Elected officials, both executive and legislative, have at times overlapping or countervailing roles, with important implications as to how bureaucracies are staffed or the amount of discretion they receive \citep{huber_legislatures_2001,gailmard_formal_2012}. Where power-sharing arrangements exist, bureaucratic appointments reflect the relative power of particular constituencies \citep{francois_how_2015}. In the context of a separation of powers, the need for alignment between executive and legislative actors may require elected officials, in particular executive leaders who have the power to appoint staff into the bureaucracy, to form coalitions with legislators \citep{groseclose_1996_buying, banks_2000_buying, gottlieb_countervailing_2019}. Where programmatic commitments are weak, coalition-formation manifests itself as transactions in the form of patronage appointments to help secure votes by legislators \citep{power_optimism_2010, amorim_neto_presidential_2000}.

From an administrative perspective, individual bureaucrats are embedded in an administrative apparatus. And a key feature that distinguishes high performing bureaucratic institutions are relatively high levels of cohesion \citet{evans_embedded_1995,evans_bureaucracy_1999}. In other words, well-functioning bureaucracies are characterized by an ability to meritocratically recruit and retain talent. This is an organizational characteristic, which goes beyond any single bureaucrat and takes into account the bureaucracy as a whole. In contrast, bureaucracies exposed to high levels of turnover, in particular when the drivers of staff turnover are political, can display low levels of cohesion and morale. \citep{kohli_state-directed_2004, akhtari_political_2015}. Patronage transactions between elected officials, in particular when these result into a wave of new political appointees, disrupt established organizational procedures, throwing management into disarray and requiring an adjustment period to the new entries.

To shed light on this complexity, over the course of two years I conducted fieldwork research in municipalities across Brazil. In interviews with local politicians, school principals and teachers, I discovered what many scholars working on the developing world already know: clientelism is pervasive, and it extends to public education with nefarious consequences \citep{stokes_brokers_2013}. However, the type of clientelism I observed was not embedded in electoral politics \citep{oliveros_making_2016} or partisan networks \citep{akhtari_political_2015,colonnelli_patronage_2017}, but the elite politics of government \citep{raile_executive_2011}. In Brazil, the exchange of public sector appointments for political favors -- a canonical example of patronage -- grants the mayor a carte blanche to enact her policy agenda.
